---
title: Using R markdown to document R programs.
---

<center>
author: A. A. El Haddi  <br>
date: Jan, 13th, 2019<br>

</center>

# Note: how to process the Rmd file. 
Use the following command line to process the Rmd file.
Best to put the following line in a bash script or Makefile and pass in the file names as arguments

This will produce both the html document and the pdf document.
I attached the pdf document because github will not show large html docs.

R -e "rmarkdown::render('$1', c('html_document', 'pdf_document'))"


In order to get the pdf documents on ubuntu (install texlive texstudio texlive-latex-extra)

 * sudo apt-get install texlive texstudio 

 * sudo apt-get install texlive-latex-extra

The code can also be found here: https://www.enduradata.com/Rcode

``` {r setup, include=TRUE}
knitr::opts_chunk$set(cache=FALSE)
```


# Include the packages that we will need

``` {r includelibs }

library(lubridate)
library(ggplot2)

```


#Loading and preprocessing the data
### Load the data 

### 1. Code for reading in the dataset and/or processing the data


``` {r input}
zipactivity <- "activity.zip"
unzipactivity <- "activity.csv"
activity <- read.csv2(unz(zipactivity, unzipactivity), header = TRUE, sep = ",",  dec = ".", fill = TRUE, na.strings = "NA")

```

## Quick check the data



``` {r checkdata }
dim(activity)
head(activity)
names(activity)

```

## Transform dates, add weekends, etc

``` {r transformsdates }

activity$date<-as.Date(activity$date, format="%Y-%m-%d")
activity$weekday <- weekdays(activity$date)
activity$dayofweek <- wday(activity$date)
activity$isweekend <- wday(activity$date) %in% c(1,7)

ssize <- dim(activity)[1]
 activity$wdaytype <- "weekday"
 activity$wcolor <- "red"

 for (i in 1:ssize) {
    if (activity[i, "isweekend"])  {
		activity[i, "wdaytype"] <- "weekend"
 		activity$wcolor <- "green"
	}
 }

head(activity)
summary(activity)

```

# Crude analysis including missing values
### 2. Histogram of the total number of steps taken each day

Ignore missing values



``` {r histwithmval }
	
stepsperday <- aggregate(activity[, "steps"], by=list(activity$date), FUN=sum, na.rm=TRUE)
colnames(stepsperday) <- c("date", "steps")
hist(stepsperday$steps, col="red", main="Number of steps per day, missing values not imputed")

```



## 3. Mean and median number of steps taken each day
## Get total steps per day

Here we did not exclude the missing values yet

* The mean total number of steps taken  is `r median(stepsperday$steps, na.rm=TRUE)`
* The median total number of steps taken is `r median(stepsperday$steps, na.rm=TRUE)`


### 4. Time series plot of the average number of steps taken

``` {r timeseriesplot}

avstepsperinterval <- aggregate(activity[, "steps"], by=list(activity$interval), FUN=mean, na.rm=TRUE)
colnames(avstepsperinterval) <- c("interval", "steps")
plot(avstepsperinterval$interval, avstepsperinterval$steps, type='l', main="Average steps per interval over all dates")
maxsteps<-max(avstepsperinterval$steps, na.rm=TRUE)


```

### 5. Maximum number of steps per interval

``` {r maxi}
 s1<-subset(avstepsperinterval, steps==maxsteps, na.rm=TRUE)
 maxi<-s1$interval
 plot(avstepsperinterval$interval, avstepsperinterval$steps, type='l', main="Average steps per interval over all dates")
 abline(v=maxi, col="red")
 slabel<-sprintf("Max steps per interval of %s occur at interval %s\n", round(maxsteps), maxi)
 text(maxi+40, maxsteps-10,slabel, col="blue")


```

The red vertical line shows that the maximum number of steps per interval occurs at `r s1$interval`


### 6. Code to describe and show a strategy for imputing missing data


## Find and treat missing values
We create another subset without missing values. We will use later to estimate missing values means.

###imputing missing values
``` {r mvvals }

# find missing values
mv <- is.na(activity$steps)

length(mv)

# set without missing values
nomvactivity<-activity[!mv,]
dim(nomvactivity)

# set with missing values
mvactivity<-activity[mv,]

print("Dimension of missing value data set")
dim(mvactivity)

``` 

Total number of missing values : `r dim(mvactivity)[1]`

```  {r imputemv }


# We impute
# For each missing value we estimate the value using non missing values for that step

 # create another instance of the activity data set but track which steps were estimated

 imactivity <- activity
 imactivity$hadmv <- FALSE
 imactivity$orgsteps <- imactivity$steps

 for (i in 1:dim(imactivity)[1]) {

	## redundant test but leave here in case we want to replace within same data set
    if (is.na(imactivity[i, "steps"]))  {

		# current interval
		 cinterval<-imactivity[i, "interval"]

		 # mean of all intervals equal to current interval without missing values
		 umv <- mean(subset(imactivity, interval == cinterval)$steps, na.rm=TRUE)

		 # Here we round up the number of steps
		 imactivity[i, "steps"] <- round(umv)
		 imactivity[i, "hadmv"] <- TRUE
	}
 }

summary(imactivity)

``` 

### 7. Histogram of the total number of steps taken each day after missing values are imputed


``` {r histwithoutmv }

# Total number of steps per day using imputed values.

imstepsperday <- aggregate(imactivity[, "steps"], by=list(imactivity$date), FUN=sum, na.rm=FALSE)
colnames(imstepsperday) <- c("date", "steps")

hist(imstepsperday$steps, col="yellow", main="Number of steps -- missing values were imputed")

```

### 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

(see transformsdates chunck where we already added weekdays, weekends etc)

``` {r facetplot }


avstepsperintervalwkday <- aggregate(imactivity[, "steps"], by=list(imactivity$interval, imactivity$wdaytype), FUN=mean, na.rm=TRUE)
colnames(avstepsperintervalwkday) <- c("interval", "wdaytype", "steps")


 ggplot(avstepsperintervalwkday, aes(x=interval, y=steps)) +
    geom_point( color="gray", size=3, alpha=.7) +
    facet_grid(. ~ wdaytype) +
    labs(x="5 minute Intervals", y="Total number of steps") +
    ggtitle("Total number of steps by 5 minutes interval by Weekend and Week day") +
    geom_line()


```


